---
title: "Limnolab_3_Phosphore"
author: "Randrianasolo Iriana"
date: "2026-03-18"
output: html_document
---

```{r}
# Morphological data

V     = 0.55 # Lake total volume [km3]
V_h   = 0.34 # Volume of the hypolimnion (< 10 m depth) [km3]
S_h   = 17   # Surface of the hypolimnion (< 10 m depth) [km2]
S_a_h = 9    # Surface of the anoxic hypolimnion (< 30 m depth) [km2]
```

•	Available datasets
- P_inflow_outflow_Murten : annual water discharges (m3/s) and average TP and PO4 concentrations (mg/m3) in lake Morat’s inflows and outflow (La Broye Canal) between 1988 and 2008 (i.e. when most remediation actions were undertaken).
- Murtensee_P: annual average concentrations in TP and PO4 (winter average) and in the epilimnion in mg/l in lake Morat 1988-2024.
-	P_sediment fluxes : estimated annual sediment burial in total phosphorus (in tP/year). These estimates were computed from dated sediment cores.
-	Murtensee_chemistry_1986_2025: complete datasets of chemical and biological monitoring in Lake Morat since 1986.


```{r}
# Load the files
P_inflow_outflow_Murten <- read.csv2("C:/Users/haren/OneDrive/Bureau/UNIL MA2/Aquatic ecosystems/LIMNOLAB_3_P-20260318/P_inflows_outflow_Murten.csv")

Murtensee_P <- read.csv2("C:/Users/haren/OneDrive/Bureau/UNIL MA2/Aquatic ecosystems/LIMNOLAB_3_P-20260318/Murtensee_P.csv") 

P_sediment_fluxes <- read.csv2("C:/Users/haren/OneDrive/Bureau/UNIL MA2/Aquatic ecosystems/LIMNOLAB_3_P-20260318/P_sediment_fluxes.csv")

#Murtensee_chemistry_1986_2025 <- read.xlsx("C:/Users/haren/OneDrive/Bureau/UNIL MA2/Aquatic ecosystems/LIMNOLAB_3_P-20260318/Murtensee_Chemistry_1986_2025.xlsx")
```

1.	Using P data from inflows estimate the relative share of the different inflows in terms of TP inputs to lake Morat in 1988 and 2008. Where were the major remediation actions undertaken? 

```{r}
library(tidyverse)
library(lubridate)

P_inflow_outflow_Murten_1988 <- P_inflow_outflow_Murten %>% filter(year==1988)
P_inflow_outflow_Murten_1988


P_inflow_outflow_Murten_2008 <- P_inflow_outflow_Murten %>% filter(year==2008)
P_inflow_outflow_Murten_2008

```
```{r}
# 1988
P_inflow_outflow_Murten_1988 <- P_inflow_outflow_Murten_1988 %>% 
  mutate(water_discharge_m3s = as.numeric(water_discharge_m3s))

P_inflow_outflow_Murten_1988 <- P_inflow_outflow_Murten_1988 %>% 
  mutate(Ptot_mgl = as.numeric(Ptot_mgl))
  
P_inflow_outflow_Murten_1988 <- P_inflow_outflow_Murten_1988 %>%
  mutate(P_annual = (water_discharge_m3s * Ptot_mgl * 24 * 3600 * 365)/10^9)

P_inflow_outflow_Murten_1988

P_in_1988 <- P_inflow_outflow_Murten_1988 %>%
  filter(type == "inflow") %>%
  mutate(total = sum(P_annual))  # Unit tonnes [g]

P_in_1988

# 2008
P_inflow_outflow_Murten_2008 <- P_inflow_outflow_Murten_2008 %>% 
  mutate(water_discharge_m3s = as.numeric(water_discharge_m3s))

P_inflow_outflow_Murten_2008 <- P_inflow_outflow_Murten_2008 %>% 
  mutate(Ptot_mgl = as.numeric(Ptot_mgl))
  
P_inflow_outflow_Murten_2008 <- P_inflow_outflow_Murten_2008 %>%
  mutate(P_annual = (water_discharge_m3s * Ptot_mgl * 24 * 3600 * 365)/10^9)

P_inflow_outflow_Murten_2008

P_in_2008 <- P_inflow_outflow_Murten_2008 %>%
  filter(type == "inflow") %>%
  mutate(total = sum(P_annual)) # Unit tonnes [g]

P_in_2008

```
```{r}
# Plot
library (ggplot2)
P_comparaison_1988_2008<- rbind(P_in_1988, P_in_2008)

P_comparaison_1988_2008 <- P_comparaison_1988_2008 %>%
  mutate(Share = P_annual/total)

ggplot(P_comparaison_1988_2008, aes(x = as.factor(year), y = P_annual, fill = as.factor(River))) +
  geom_col() + 
  labs(title = "Comparaison de la charge de Phosphore : 1988 vs 2008",
       x = "Year",
       y = "Annual P (tP/y)",
       fill = "Inflow source") +
  theme_minimal()
```
2.	From 2008 onwards, we consider P inflows to be constant. The Ptot concentration in the lake are stable (i.e. we’ve reached steady state). What does that mean for the balance in P of the lake.
Steady state >=> dP/dt = 0

It means that P_in - P_out -P_NetSed = 0

```{r}
P_sediment_fluxes_2008 <- P_sediment_fluxes %>%
  filter(year == 2008)

P_in  = 39.15923  # [tP/y]
P_out = 4.5222624 # [tP/y]
P_NetSed = 41	


P_in-P_out-P_NetSed # = 6.4 which is the internal load or due to P diffusion
```

5. 	Sediment core incubations showed a release of 0.23 mmol/m2/d of P in anoxic conditions. Considering that over 100 days per year, the whole hypolimnion gets anoxic up to the 10m depth (estimated sediment surface area of 9 km2), what is the amount of P released annually by the sediment? Conclude on the current P budget of Lake Morat.

```{r}
release_rate <- 0.23       # [mmol/m2/d]
duration_days <- 100       # Days of anoxia per year 
molar_mass_P <- 30.97      # g/mol

S_a_h_m2 <- S_a_h * 1e6  

total_mmol <- release_rate * S_a_h_m2 * duration_days
total_grams <- total_mmol * (molar_mass_P / 1000)
total_tons <- total_grams / 1e6

cat("Annual P released by sediment:", total_tons, "tP/yr")
```
7. Assess how the Schiffnen connection will affect the final lake P concentrations
```{r}
# Beta
Murtensee_P_2008 <- Murtensee_P %>% 
  filter(Year == 2008) %>%
  mutate(P_winter = as.numeric(TP_winter_mgl), 
         P_epi = as.numeric(TP_epi_mgl))

beta <- Murtensee_P_2008$TP_winter_mgl / Murtensee_P_2008$TP_epi_mgl
cat("Parameter beta:", beta, "\n")


# Sigma 
P_mass_lake_2008 <- Murtensee_P_2008$TP_winter_mgl * V * 1000 # [tP]

sigma <- P_NetSed / P_mass_lake_2008
cat("Parameter sigma (yr-1):", sigma, "\n")


#Phosphorus  After Connection
Q_pump_m3s <- 100
P_schiff_mg_m3 <- 40 
P_schiff_load_t <- (Q_pump_m3s * P_schiff_mg_m3 * 24 * 3600 * 365) / 10^6 # [tP/yr]

# New Inputs
Pin_new <- P_in + P_schiff_load_t

# Current residence time 
Q_total_2008 <- sum(P_in_2008$water_discharge_m3s)
tau_old <- (V * 10^9) / (Q_total_2008 * 24 * 3600 * 365) 

# Residence time after connection
tau_new <- (V * 10^9) / ((Q_total_2008 + Q_pump_m3s) * 24 * 3600 * 365) 

# Steady State Equilibrium Concentration [P] 
P_eq_new <- (Pin_new * 10^9) / ((V * 10^9) * (sigma + (beta / tau_new)))

cat("New Equilibrium P Concentration (mg/m3):", P_eq_new, "\n")

#  P Retention Rate
Pout_old <- (beta / tau_old) * (P_conc_mix_2008 * 1000) * V # [tP/yr]
retention_old <- (P_in - Pout_old) / P_in

Pout_new <- (beta / tau_new) * (P_eq_new / 1000) * (V * 10^9) / 10^6 # [tP/yr]
retention_new <- (Pin_new - Pout_new) / Pin_new

cat("Residence time before", tau_old*365 , "d \n")
cat("Residence time after", tau_new*365 , "d \n")
cat("Retention Rate Before:", retention_old * 100, "%\n")
cat("Retention Rate After:", retention_new * 100, "%\n")
```





